---
title: "ZebraBox Analysis Report"
output: html_document
params:
  metadata_file: NULL
  data_file: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(bslib)
library(vroom)
library(broom)
library(dplyr)
library(ggplot2)
library(stringr)
library(writexl)
library(DT)
library(plotly)
library(tidyr)
library(purrr)
library(tibble)
library(ggbeeswarm)

```

## 


```{r pressure, echo=FALSE}

# reading in the data

control_group <- "WT"

metadata <- readxl::read_excel(params$metadata_file, skip = 1) %>%
  mutate(box_used = as.character(box_used),
         box_used = case_when(box_used == '1' ~ 'LocA',
                              box_used == '2' ~ 'LocB',
                              TRUE ~ NA_character_))

zebrabox_data <- read.delim(params$data_file, fileEncoding = "UTF-16LE", stringsAsFactors = FALSE) %>%
  mutate(time_min = start / 60,
         experiment_time_of_day = '',
         period2 = ifelse(nchar(time_min) == 1, 1,
                          as.integer(str_extract(time_min, '^[0-9]')) + 1),
         plate = str_extract(location, 'Loc[AB]'),
         light = ifelse(period2 %% 2 != 0 & period2 != 1, 'dark', 'light')) %>%
  separate(aname, into = c('well', 'name'), sep = '_') %>%
  left_join(metadata, by = join_by(well, plate == box_used)) %>%
  select(plate, treatment, well, time_min, light, period,
         datatype, activity = actinteg, everything()) %>%
  filter(!is.na(treatment), timebinid == 1)

```

### All Fish by Minute Plot
```{r, fig.width = 16}

zebrabox_data %>% 
  filter(startreason == 'Beginning of period', !is.na(activity)) %>%
  filter(ifelse(str_detect(plate, 'LocB'), plate == 'LocB', plate == 'LocA')) %>%
  filter(activity == max(activity)) %>%
  select(activity) %>%
  deframe() -> qc_all_act_max

# zebrabox all fish
zebrabox_data %>% 
  filter(startreason == 'Beginning of period') %>%
ggplot(aes(x = time_min, y = activity, color = well, group = well)) +
  geom_rect(aes(fill = light, xmin = time_min,
                 xmax = time_min + 1, ymin = 0, ymax = qc_all_act_max + 100), 
            color = NA) +
  scale_fill_manual(values = c('gray80', 'white')) +
  geom_vline(xintercept = seq(0, max(zebrabox_data$time_min), 10), linetype = 'dashed', color = 'gray60') +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = seq(0, max(zebrabox_data$time_min), 10)) +
  facet_wrap(~ plate, ncol = 1) + 
  coord_cartesian(ylim = c(0, qc_all_act_max)) +
  labs(x = 'Time (min)', y = 'Activity', title = 'Zebrabox Individual Fish') +
  theme_classic(base_size = 16) -> zebrabox_individual
ggplotly(zebrabox_individual)
```

### Condition Summary Plot
#### An interactive plot showing summarized activity levels over time for different experimental conditions, with labels such as (dark, 1)/(light, 1) indicating the light/dark condition and colors indicating the users’ specified treatment.
```{r, fig.width = 16}

### calculate average activity by treatment
zebrabox_data %>%
  filter(startreason == 'Beginning of period') %>% 
  group_by(plate, treatment, light, time_min) %>%
  summarize(mean_activity = mean(activity, na.rm = T)) %>%
  ungroup() -> qc_condition_summary

qc_condition_summary %>%
  filter(ifelse(str_detect(plate, 'LocB'), plate == 'LocB', plate == 'LocA')) %>%
  filter(mean_activity == max(mean_activity)) %>%
  select(mean_activity) %>%
  deframe() -> qc_condition_max
  
### plot an interactive graph of summarized activity by treatment
qc_condition_summary %>%
ggplot(aes(x = time_min, y = mean_activity, color = treatment, group = treatment)) +
  geom_rect(aes(fill = light, xmin = time_min,
                 xmax = time_min + 1, ymin = 0, ymax = qc_condition_max + 100), color = NA) +
  scale_fill_manual(values = c('gray80', 'white')) +
  geom_vline(xintercept = seq(0, max(zebrabox_data$time_min), 10), linetype = 'dashed', color = 'gray60') +
  geom_point() +
  geom_line() +
  facet_wrap(~ plate) +
  coord_cartesian(ylim = c(0, qc_condition_max)) +
  labs(x = 'Time (min)', y = 'Activity', title = 'Zebrabox Summarized Conditions') +
  theme_classic(base_size = 16) -> zebrabox_summarized
ggplotly(zebrabox_summarized)
```

### Plate Map 
#### An interactive plate map that visualizes activity levels in each plate well.
```{r}
zebrabox_data %>%
  separate(well, into = c('well_row', 'well_column'), sep = 1) %>%
  filter(datatype == 'QuantizationSum', period2 == 3) %>%
ggplot(aes(x = well_column, y = well_row, text = treatment,
           color = activity)) +
  geom_point(size = 4) +
  scale_color_continuous(limits = c(0, 7000)) +
  scale_x_discrete(position = "top")  +
  scale_y_discrete(limits = rev) +
  facet_wrap(~ plate, ncol = 2) +
  labs(x = NULL, y = NULL, color = 'Activity\n1st Dark\nPeriod') +
  theme_minimal() -> plate_map

ggplotly(plate_map)
```

### Mean Activity Plot

```{r}
zebrabox_data %>%
  filter(datatype == 'QuantizationSum', !is.na(activity)) %>%
  group_by(treatment, light, well) %>%
  summarize(mean_activity = mean(activity)) %>%
  ungroup() %>%
  filter(light == 'dark') %>%
ggplot(aes(x = treatment, y = mean_activity)) +
  ggbeeswarm::geom_quasirandom() +
  geom_boxplot(alpha = 0) +
  labs(x = 'Condition', 
       y = 'Mean Activity Dark Cycles',
       title = 'Unnormalized Activity') +
  theme_bw()
```

<!-- ### Percent Change Plot -->
<!-- ```{r} -->
<!-- zebrabox_data %>% -->
<!--   filter(datatype == 'QuantizationSum', !is.na(activity), -->
<!--          strain == control_group, light == 'dark') %>% -->
<!--   summarize(median(activity)) %>% -->
<!--   deframe() -> median_activity -->


<!-- zebrabox_data %>% -->
<!--   filter(datatype == 'QuantizationSum', !is.na(activity)) %>% -->
<!--   mutate(percent_change = ((activity - median_activity) / median_activity) * 100) %>% -->
<!--   group_by(treatment, light, well) %>% -->
<!--   summarize(mean_activity = mean(percent_change)) %>% -->
<!--   ungroup() %>% -->
<!--   filter(light == 'dark') %>% -->
<!-- ggplot(aes(x = treatment, y = mean_activity)) + -->
<!--   ggbeeswarm::geom_quasirandom() + -->
<!--   geom_boxplot(alpha = 0) + -->
<!--   labs(x = 'Condition',  -->
<!--        y = 'Average Percent Change From Control Median\nin Dark Cycles', -->
<!--        title = 'Percent Change Activity') + -->
<!--   theme_bw() -->

<!-- zebrabox_data %>% -->
<!--   filter(datatype == 'QuantizationSum', !is.na(activity)) %>% -->
<!--   mutate(percent_change = ((activity - median_activity) / median_activity) * 100) %>% -->
<!--   group_by(treatment, light, well) %>% -->
<!--   summarize(mean_activity = mean(percent_change)) %>% -->
<!--   ungroup() %>% -->
<!--   filter(light == 'dark') %>% -->
<!--   rename(mean_percent_change = mean_activity) %>% -->
<!-- DT::datatable(extensions = 'Buttons', -->
<!--                 options = list(dom = 'Blfrtip', -->
<!--                               buttons = c('copy', 'csv', 'excel'), -->
<!--                               lengthMenu = list(c(10, 25, 50, -1), -->
<!--                                                 c(10, 25, 50, "All")))) -->
<!-- ``` -->



### Activity by Period
```{r, fig.width = 12, fig.height = 4}
zebrabox_data %>%
  filter(datatype == 'QuantizationSum', !is.na(activity)) %>%
  mutate(period3 = ifelse(period2 %in% 1:2, 
                          'acclimation', as.character(period2)),
         period3 = factor(period3, levels = c('acclimation', '3', '4', '5', 
                                              '6', '7', '8', '9', '10'))) %>%
  group_by(period3, treatment, light, well) %>%
  summarize(mean_activity = mean(activity)) %>%
  ungroup() %>%
ggplot(aes(x = period3, y = mean_activity, color = treatment)) +
  annotate("rect", xmin = 1.5, xmax = 2.5, ymin = 0, ymax = Inf, fill = 'gray80') +
  annotate("rect", xmin = 3.5, xmax = 4.5, ymin = 0, ymax = Inf, fill = 'gray80') +
  annotate("rect", xmin = 5.5, xmax = 6.5, ymin = 0, ymax = Inf, fill = 'gray80') +
  annotate("rect", xmin = 7.5, xmax = 8.5, ymin = 0, ymax = Inf, fill = 'gray80') +
  geom_vline(xintercept = seq(1.5, 8.5, 1), linetype = 'dashed', color = 'gray60') +
  geom_boxplot(alpha = 0) +
  # ggbeeswarm::geom_quasirandom(position = position_dodge(width = .75)) +
   geom_point(position = position_dodge(width = .75)) +
  labs(x = 'Cycle', 
       y = 'Mean by Dark Cycle',
       title = 'Unnormalized Activity') +
  theme_bw() -> act_by_period

act_by_period
```


### Statistical Summary
#### An interactive table providing the results of the statistical tests.
```{r}
zebrabox_data %>%
  filter(datatype == 'QuantizationSum', !is.na(activity)) %>%
  mutate(period3 = ifelse(period2 %in% 1:2, 
                          'acclimation', as.character(period2)),
         period3 = factor(period3, levels = c('acclimation', '3', '4', '5', 
                                              '6', '7', '8', '9'))) %>%
  group_by(period3, treatment, light, well) %>%
  summarize(mean_activity = mean(activity)) %>%
  ungroup() %>%
  filter(light == 'dark') %>%
  group_by(period3) %>%
  nest() %>%
  ungroup() %>%
  mutate(test = map(data, ~ TukeyHSD(aov(mean_activity ~ treatment, data = .))),
         class = map(test, ~ class(.))) %>%
  unnest(c(class)) %>%
  filter(class != 'try-error') %>%
  mutate(test = map(test, ~ broom::tidy(.))) %>%
  unnest(c(test)) %>%
  select(period = period3, 
         groups_tested = contrast,
         difference_means = estimate,
         conf.low, conf.high, adj.p.value) %>%
  mutate(difference_means = round(difference_means),
         conf.low = round(conf.low),
         conf.high = round(conf.high),
         adj.p.value = round(adj.p.value, 4)) %>%
  mutate(significant = ifelse(adj.p.value < 0.05, 'significant', 'not significant')) %>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                              buttons = c('copy', 'csv', 'excel'),
                              lengthMenu = list(c(10, 25, 50, -1),
                                                c(10, 25, 50, "All"))))
```

### Cleaned Data Table
```{r, fig.width = 16}

zebrabox_data %>%
DT::datatable(extensions = 'Buttons',
               options = list(dom = 'Blfrtip',
                              buttons = c('copy', 'csv', 'excel'),
                              lengthMenu = list(c(10, 25, 50, -1),
                                                c(10, 25, 50, "All"))),
               caption = htmltools::tags$caption(style = 'caption-side: top; text-align: left; color:black;  font-size:200% ;',
                                                   'Zebrabox'))

```

```{r}
zebrabox_data %>%
  filter(datatype == 'QuantizationSum', !is.na(activity)) %>%
  group_by(treatment, light, period2, well) %>%
  mutate(index = row_number()) %>%
  nest() %>%
  ungroup() %>%
  mutate(period_peak = map(data, ~ as_tibble(pracma::findpeaks(.$activity, npeaks = 1)))) %>%
  unnest(c(period_peak)) %>%
  rename(peak_height = V1, peak_index = V2, curve_start_index = V3,
         curve_end_index = V4) %>%
  unnest(c(data)) %>%
  mutate(max_peak = ifelse(index == peak_index, 'max', NA_character_)) -> zebrabox_data_w_peaks
```

  
### Starting Slope

```{r}
# zebrabox
zebrabox_data_w_peaks %>% #distinct(period) %>% mutate(period2 = str_remove(period, '^0+:')
  group_by(treatment, well, light, period2) %>%
  filter(index <= peak_index) %>% #filter(well == 'c1-004') %>% tail(1) %>% nest() %>% ungroup() %>% mutate(test = map(data, ~ .$activity)) %>% unnest(c(test))
  nest() %>%
  ungroup() %>%
  mutate(len = map(data, ~nrow(.))) %>%
  unnest(c(len)) %>%
  mutate(period_start_slope = case_when(len == 1 ~ map(data, ~ .$activity), # maybe return 0 or NA?
                                        len == 2 ~ map(data, ~ as.numeric(dist(select(.,
                                                                                      activity, time_min),
                                                                               method = 'euclidean'))),
                                        TRUE ~ map(data, ~ filter(broom::tidy(lm(activity ~ time_min, data = .)),
                                                                  term == 'time_min')$estimate))) %>%
  unnest(c(period_start_slope)) %>%
  distinct(treatment, period2, light, well, period_start_slope) %>%
  filter(light == 'dark') %>%
ggplot(aes(x = period2, y = period_start_slope, color = treatment, group = well)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = c(3, 5, 7, 9)) +
  labs(x = 'Condition',
       y = 'Starting Slope Dark Cycles',
       title = 'Zebrabox') +
  facet_wrap(~ treatment) +
  theme_bw()
```

### Ending Slope

```{r}
### just used a linear model
# zebrabox
zebrabox_data_w_peaks %>%
  group_by(treatment, well, light, period2) %>%
  filter(index >= peak_index) %>% #filter(well == 'c1-004') %>% tail(1) %>% nest() %>% ungroup() %>% mutate(test = map(data, ~ .$activity)) %>% unnest(c(test))
  nest() %>%
  ungroup() %>%
  mutate(len = map(data, ~nrow(.))) %>%
  unnest(c(len)) %>%
  mutate(period_start_slope = case_when(len == 1 ~ map(data, ~ .$activity),
                                        len == 2 ~ map(data, ~ as.numeric(dist(select(.,
                                                                                      activity, time_min),
                                                                               method = 'euclidean'))),
                                        TRUE ~ map(data, ~ filter(broom::tidy(lm(activity ~ time_min, data = .)),
                                                                  term == 'time_min')$estimate))) %>%
  unnest(c(period_start_slope)) %>%
  distinct(treatment, period2, light, well, period_start_slope) %>%
  filter(light == 'dark') %>%
ggplot(aes(x = period2, y = period_start_slope, color = treatment, group = well)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = c(3, 5, 7, 9)) +
  labs(x = 'Condition',
       y = 'Ending Slope Dark Cycles',
       title = 'Zebrabox') +
  facet_wrap(~ treatment) +
  theme_bw()
```
